Fail Faster

### 2025-03-17 paper/paper.tex
reordering

\section{Background}

\jwc{In this paper, we'll use OCaml to illustrate the ideas, but the concepts
are portable to other languages: see Section~\ref{subsection:other-langs}}

\subsection{Property-Based Testing}
\jwc{Usual spiel about pbt and generators: here's a monadic generic generator library (but we're using BQ to illustrage), it has return and bind, this is what they're for.}
\jwc{You often want generators to generate for sparse precoditions so you can exercise a property reasonably well}

\begin{lstlisting}
module Bq = struct
  type 'a t = int -> SR.t -> 'a

  let return (x : 'a) : 'a t = fun size seed -> x

  let bind (g : 'a t) (f : 'a -> 'b t) : 'b t =
    fun size seed ->
      let x = g size seed in
      (f x) size seed

  let gen_int (lo : int) (hi : int) : int t =
    fun size seed -> SR.int seed lo hi
end
\end{lstlisting}

@@

\subsection{Multi-Stage Programming}
\label{subsection:msp}

\jwc{I think this section should actually go later.}

\subsubsection{Staging in eDSL Construction}
One of the primary uses of staging is in embedded DSL construction \cite{sheard99, lms, trattdsl}.
While embedded DSLs are great \jwc{words}, they have well-known drawback: 
the functional abstractions used to build eDSLs prevent compilers from generating efficient machine code \cn.
In the literature, this effect is known as ``abstraction overhead'' \cite{carrette05, moller20}: the layers of abstraction that 
make eDSLs so usable are precisely what prevents them from being fast.
In light of this issue, staging is often used as a lightweight compiler for DSLs. The compile-time evaluation stage
transforms the DSL code, eliminating abstractions to produce code that the host language compiler can generate fast machine code for.
\jwc{This previous paragraph is peak ``shitty first draft''.}
This recipe has been used to great effect across domains to stage eDSLs for stream processing \cite{moller20,strymonas}, parser combinators \cite{sspc, staged-parsers, krishnaswami19, flap},
and query processing \cite{rhyme}.

\subsection{Other Languages and Libraries}
\label{subsection:other-langs}

\section{Sources of Inefficiency in Monadic Generator Libraries}
\label{section:motiv}
\jwc{In this section we answer: ``Why are monadic generator libraries slow?''}

@@

\subsection{Monadic DSL Abstraction Overhead}
Just how large is the abstraction overhead of monadic generator DSLs, and where does it come from?
Figure~\ref{fig:bq-internals} shows the internals of (a simplified version of) \bq{}.
A generator \texttt{'a Bq.t} is just a function of type~\texttt{int -> SR.t -> 'a},
taking an \texttt{int} representing the current size parameter and a random seed \texttt{SR.t},
and returning a generated value \texttt{'a}. It is an invariant of the library that function of this type are
deterministic: for a fixed size and seed, it will return the same value. All of the randomness in testing comes from varying the initial seed.
The monad functions \texttt{return} and \texttt{bind} are defined in the usual way
for an instance of the reader monad \cn{}. \texttt{Return x} ignores the size and seed and just returns
\texttt{x}, while \texttt{bind g k} first runs \texttt{g} and then passes the generated value to the continuation \texttt{k}.
\bq{} uses a mutable seed, so the \texttt{random} passed to \texttt{(k a)} will be different from the one passed to \texttt{g}.
The \texttt{gen\_int} combinator simply calls out to the randomness library \texttt{Splittable\_random}, aliased as \texttt{SR} here.
\jwc{Should we do this?}

@@

\subsubsection{Overhead of Other Combinators}

@@

% \subsubsection{Function call overhead}
\subsection{Choice of Randomness Library}
\jwc{We really need to have explained the effect ordering equivalence aspect of the evaluation before here.}

@@

@@

\section{Conclusion \& Future Work}
\section{Related Work}
\jwc{A generator is a parser of randomness, and people have implemented lots of staged parser libraries!}

>>>>

\section{Sources of Inefficiency in Monadic Generator Libraries}
\label{section:motiv}
\jwc{In this section we answer: ``What are monadic generator libraries, and why are they slow?''}

\subsection{Background: Monadic Generator DSLs}
\jwc{Usual spiel about pbt and generators: here's a monadic generic generator library (but we're using BQ to illustrage), it has return and bind, this is what they're for.}
\jwc{You often want generators to generate for sparse precoditions so you can exercise a property reasonably well}

\begin{lstlisting}
module Bq : sig
  type 'a t
  val return : 'a -> 'a t
  val bind : 'a t -> ('a -> 'b t) -> 'b t
  val gen_int : lo:int -> hi:int -> int t
end
\end{lstlisting}

@@

@@

\subsection{Abstraction Overhead of Generator DSLs}
Just how large is the abstraction overhead of monadic generator DSLs, and where does it come from?
Figure~\ref{fig:bq-internals} shows the internals of (a simplified version of) \bq{}.
A generator \texttt{'a Bq.t} is just a function of type~\texttt{int -> SR.t -> 'a},
taking an \texttt{int} representing the current size parameter and a random seed \texttt{SR.t},
and returning a generated value \texttt{'a}. It is an invariant of the library that function of this type are
deterministic: for a fixed size and seed, it will return the same value. All of the randomness in testing comes from varying the initial seed.
The monad functions \texttt{return} and \texttt{bind} are defined in the usual way
for an instance of the reader monad \cn{}. \texttt{Return x} ignores the size and seed and just returns
\texttt{x}, while \texttt{bind g k} first runs \texttt{g} and then passes the generated value to the continuation \texttt{k}.
\bq{} uses a mutable seed, so the \texttt{random} passed to \texttt{(k a)} will be different from the one passed to \texttt{g}.
The \texttt{gen\_int} combinator simply calls out to the randomness library \texttt{Splittable\_random}, aliased as \texttt{SR} here.
\jwc{Should we do this?}

@@

\subsubsection{Overhead of Choice Combinators}

@@

% \subsubsection{Function call overhead}
\subsection{Inefficient Randomness Libraries}
\jwc{We really need to have explained the effect ordering equivalence aspect of the evaluation before here.}

@@

\subsection{Background: Multi-Stage Programming}
\label{subsection:msp}

\subsubsection{Uses of Staging in eDSL Construction}
One of the primary uses of staging is in embedded DSL construction \cite{sheard99, lms, trattdsl}.
While embedded DSLs are great \jwc{words}, they have well-known drawback: 
the functional abstractions used to build eDSLs prevent compilers from generating efficient machine code \cn.
As alluded to previously \jwc{This sentence here because of the section-ordering transpose... may need to mention this earlier.}, this effect is known as ``abstraction overhead'' \cite{carrette05, moller20}: the layers of abstraction that 
make eDSLs so usable are precisely what prevents them from being fast.
\jwc{Something to the effect of ``the issues we noticed in BQ and SC exist across many eDSLs''}
In light of this issue, staging is often used as a lightweight compiler for DSLs. The compile-time evaluation stage
transforms the DSL code, eliminating abstractions to produce code that the host language compiler can generate fast machine code for.
\jwc{This previous paragraph is peak ``shitty first draft''.}
This recipe has been used to great effect across domains to stage eDSLs for stream processing \cite{moller20,strymonas}, parser combinators \cite{sspc, staged-parsers, krishnaswami19, flap},
and query processing \cite{rhyme}.

@@

\section{Other Languages and Libraries}
\label{subsection:other-langs}

\section{Related Work}
\jwc{A generator is a parser of randomness, and people have implemented lots of staged parser libraries!}

\section{Conclusion \& Future Work}
